---
title: "R Consortium"
author: "Stephen Uong & Christina Mehranbod"
date: "2024-05-22"
output: html_document
---

##Package Installation

First, let's install all necessary packages for this exercise
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

install.packages("tidycensus")
install.packages("tidyverse")

# stephen added this
#christina added this 
#added again

#HI STEVO ITS CHRISTINA 

```

## Census API Key

First, you must sign up to get a Census API key: https://api.census.gov/data/key_signup.html

```{r census_api}

library(tidycensus)

census_api_key("INSERT API KEY", install = TRUE)

#this is christina's but delete later

census_api_key("0d1515194f46f009f34f94afcbf045315abdbfbd", install = TRUE)


```

## Explore Census Variables

You can get a list of variables using tidycensus. 

You can get a list of variables from the Dicennial Census or the American Community Survey. Since we are interested in the year 2021, let's use the American Community Survey. We will choose the 5-year ACS estimates because they tend to be more stable than 1 year estimates. 

```{r variables, echo=FALSE}

variables_2021 <- load_variables(2021, "acs5", cache = TRUE)

View(variables_2021)

library(knitr)


```


#Using get_acs()
To download ACS data, we need three pieces of information. 

First, what do we want our geography to be? What's our unit of analysis? 
For this we have decided to use the Census tracts of New York State. See a full list of available geographies here: https://walker-data.com/tidycensus/articles/basic-usage.html#geography-in-tidycensus

Second, what variables do we want? We can pull those directly in using the same command line. We can even give it our own titles. 

Third, what year of data do we want? There's room for that too! 

```{r acs, echo=FALSE}

#what variables on the census tract level? 
library(tidyverse)
variables_tract <- variables_2021 %>%
  filter(geography=="tract")

#let's use median income in the past 12 months B07011_001 and educational attainment (B16010_001, B16010_002, B16010_015, B16010_028, B16010_041)

#@STEVO: FOR THE LEVEL OF CENSUS TRACT, YOU NEED TO PUT A STATE - SHOULD WE JUST PICK NY FOR THIS EXAMPLE?
acs_data <- get_acs(geography = "tract",
                    variables = c(hhincome = "B07011_001",
                                  total_education = "B16010_001", 
                                  education_lessthanhs= "B16010_002", 
                                  education_hs = "B16010_015", 
                                  education_somecollege = "B16010_028", 
                                  education_bachelors = "B16010_041"),
                    state="NY",
                    year=2021,
                    output = "wide")

#this command brings in the data as a dataframe, which means there is no geometry attached
class(acs_data)

#Explore data

#put histograms here or something like that

#to also bring in geometries, modify the command to bring in geometries
acs_data_geo <- get_acs(geography = "tract",
                    variables = c(hhincome = "B07011_001",
                                  total_education = "B16010_001", 
                                  education_lessthanhs= "B16010_002", 
                                  education_hs = "B16010_015", 
                                  education_somecollege = "B16010_028", 
                                  education_bachelors = "B16010_041"),
                    state="NY",
                    year=2021,
                    output = "wide", 
                    geometry=TRUE)

#using base R to plot

plot(acs_data_geo)

#this isn't the best way to plot but just to show that we have a map!


```

##Historical Data Import

IPUMS (put what is stands for) is a data source for... 

```{r ipums, echo=TRUE}



```


##PLACES Import
```{r places, echo=TRUE}



```

##Beautiful Maps Using ggplot()

```{r mapping, echo=FALSE}

variables_2021 <- load_variables(2021, "acs5", cache = TRUE)

View(variables_2021)

library(knitr)


```

##Customizing Maps: Labels, Arrows, Colors and More!

```{r mapping_custom, echo=FALSE}



```
